### 2018-03-29 Datelife_paper1.Rmd
paper improvement

<!--INTRODUCTION-->
Divergence time of lineages<!--named biological entities--> <!--(taxon ages)--> constitutes in many ways the main knowledge necessary for evolutionary understanding.
<!--Along with phylogenetic relationships, it is the most important piece of information allowing to test alternative evolutionary hypothesis in many areas of research, from developmental to ecological questions.
Allong with migration rates and patterns, it is fundamental to test biogeographical hypothesis-->
Coupled to species number and distribution, it is the basic information for the study of diversification processes, such as the tempo and mode of speciation and extinction, crucial for the understanding of how biodiversity patterns are shaped across space and time [@Morlon2014].

When organisms are preserved in a fossil form, a time frame of taxon origin can be obtained directly from the age of rock strata.
However, not all organisms fossilize well or at all.
Fossilization success alone is highly circumstancial, and varies depending on a number of parameters including the nature of the habitat, population size, species range breadth and physical characteristics of the organism. <!--fossil recovery is also highly circumstancial-->
Thus, relying only on the fossil record to obtain a time frame of lineage divergence<!--taxon ages--> for all life is<!--, alas,--> not possible.
Relative rate of DNA or aminoacid substitution constitutes another important source of information on lineage divergence. <!--describe the two components of rel subst rate?-->
It can be obtained from hypothesis of character homology (alignments) when reconstructing phylogenetic relationships. Molecular dating techniques use external data such as absolute time calibrations (e.g., fossils, geologic events) or absolute substitution rates <!--birds--> to generate dated phylogenies (chronograms) which contain information on absolute times of node divergence and taxon ages. <!--With molecular dating methods, an external calibration point is still necessary to estimate the absolute time of origin. Hence, the accumulation of data on fossils for calibration is also continual-->
<!--This method has been particularly useful to obtain time frame of origin of organisms with poor or no fossil record, but is also largely used for organisms with rich fossil records-->
<!--Models allow to take into account the nature of the fossil record: Given the characteristics of the fossilization process, the quality of dates inferred directly from it is not always considered the best.-->

@@

<!--What led to datelife development?
Describe public and research necessities covered by datelife
It is important to use available data on time frame of lineage origin:
To know the state of dating for a group of interest:
What range of estimated ages exist already? Are fossil and molecular
time frames coherent? (e.g., [@Magallon2015c]).
To construct a time tree of life.
For science communication, improve scientific discussions, time-framing
other events of importance in other research areas.
Comparative method: autocorrelation (phylogenetic relatedeness) corrections, phylogenetic signal
-->
Despite its importance, analytical tools to summarize available information on taxon ages are still lacking.
Data might not have been exploited because:
Data is in different repositories and formats;
Lineage names are different among studies and difficult to reconcile;
Taxonomy is also different among studies and difficult to reconcile.
<!--Data curation is necessary at some point (is this always true?)-->
Also, data curation is an important part of any biological study. The research community considers it as an important or even crucial step before data analysis. Hence, automated processes for large data analysis are frequently received with skepticism.
Recent work on this area (e.g.,, supersmart<!--and, which others?-->) aims to:
Generate new dates using all available DNA sequence information;
Perform one global analysis using all available information;
Problems or downsides: This might be time consuming for large groups and a lot of data curation is still necessary.
<!--General issues with dating techniques:-->
Choosing correct fossils for calibration requires a lot of expertise and knowledge on the group.
Incorrect use of fossils can generate severe bias in dating results [@Sauquet2012c].
DateLife palliates this by using only information available from already published studies, which are ideally constructed using robust information, such as sequence data and curated fossil calibrations.
<!--Data curation is largely based on taxonomic knowledge; or at least on the interplay between taxonomy and phylogenetics-->
<!--Importance of datelife:-->
DateLife can summarize this information in several formats that can be easily inspected by users.
This  allows rapidly obtaining a time frame of taxon divergence for a wide number of taxa.
DateLife can also generate chronograms for taxa with little available information, by using the available data as calibration points.
DateLife is the main service for scaling phylogenetic trees in Phylotastic! system [@Stoltzfus2013] <!--to construct phylogenetic trees on the fly [http://phylotastic.org/].-->
It can be used through an R package <link to documentation>, a web interface <http://www.datelife.org/query/> and an API <!--(still not up, right?)-->

@@

DateLife also stores information on input taxon presence/absence across subset chronograms. Users can choose to add ages of missing taxa to subset chronograms in different ways, depending on the amount of knowledge or how much they want to be involved in the steps of the addition process.
If users have no access to biological information (i.e., a character, DNA or protein matrix), missing taxa can be added to any chronogram simply at random, or by following taxonomic or phylogenetic knowledge from expert sources. There are a wide number of open reference taxonomies available, such as the Catalogue of Life [@Roskob2017] or the NCBI taxonomy database [@Federhen2012]. Expert phylogenies to be used as topological constraint or backbone (a tree with no branch lengths) can also be obtained from a number of public repositories, such as OToL [@Hinchliff2015], TreeBASE [@Piel2002] and Dryad <https://www.datadryad.org//>.
At the moment, DateLife only uses OToL's synthetic tree and reference taxonomy as expert knowledge to automatically add missing taxa to chronograms. Alternatively, users can input a reference taxonomy or topological constraint of their choosing or making. <!-- at some point we will implement the possibility for users to choose different reference taxonomies automatically-->
If OToL's synthetic tree is not satisfactorily resolved for the taxa of interest, DateLife can construct a sequence data matrix from DNA markers available from the Barcode of Life Database (BOLD; @Ratnasingham2007), to attempt to further resolve polytomies. It will follow OToL's synthetic tree as backbone. <!--This molecular tree can then be dated with chronosMPL from the ape package, but is it really needed?? To estimate node ages chronosMPL uses mean path length method from Britton et al. 2002-->
<!--It can rapidly construct trees with branch lengths from
a set of lineages with sequence data from the Barcode Of Life Database (BOLD, <http://www.boldsystems.org/>), using the synthetic OToL (<https://tree.opentreeoflife.org>) as backbone, and to obtain divergence dates.
It allows direct comparison of dates obtained with different markers
available in BOLD (in plants and fungi in particular).-->
To use information from a topological constraint, DateLife calls the congruification method described in [@Eastman2013] to find shared nodes between trees (congruent nodes). It then fixes their ages, and add ages to remaining nodes with a dating method that can be specified by the user. If users have access to biological data, thay can input a tree with branch lengths proportional to relative substitution rates as topological constraint. In this case, age data from congruent nodes will be used as calibration points.
Age data from several chronograms can be combined and congruified to be used as calibration points in a single analysis.

Several dating methods are implemented in DateLife.
Branch Length Adjuster (BLADJ) is a simple algorithm to distribute ages of undated nodes evenly, which minimizes age variance in the chronogram [@Webb2008].
DateLife implements BLADJ from the development R version of phylocom's R package [@Webb2008], phylocomr <https://github.com/ropensci/phylocomr>. It can only be used when there is a topological constraint with no branch lengths.
PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees. It is also called through R. <!--It could be used with a reference taxonomy, if we provide a tree with politomies-->
treePL, is a semi-parametric, rate-smoothing, penalized likelihood dating method [@Smith2012]. It is called through R.
MrBayes program [@Huelsenbeck2001; @Ronquist2003] can be used when adding taxa at random, following a reference taxonomy or a topological constraint. It draws ages from a pure birth model, as implemented by Jetz and collaborators [-@Jetz2012].
DateLife calls MrBayes trough an R function.

@@

<!--Potential applications shown here-->Here, we have shown that DateLife allows an easy and fast obtention of all publicly available information on taxon ages, which can be used to generate new data. <!--Potential applications not shown here-->This information can be used to account for the effect of phylogenetic signal in studies of trait evolution; to explore potential speciation and extinction dynamics of interest within a clade; to obtain a time frame of biogeographical events; for science communication and outreach, amongst others.
<!--Why you should use it-->Compared to similar platforms such as time tree of life and supermart, it offers several advantages.
It is fast;
source data is completely open;
it requires no expert biological knowledge for any of their functionalities;
it allows exploration of alternative taxonomic and phylogenetic schemes;
it allows rapid exploration of the effect of alternative divergence time hypothesis;
it allows rapid synthesis in a number of different formats; <!-- we should add graphical options too-->
it facilitates reproducibility of analyses;

>>>>

<!--INTRODUCTION-->
Divergence time of lineages<!--named biological entities, taxon ages?--> constitutes in many ways the main knowledge necessary for evolutionary understanding.
<!--Along with phylogenetic relationships, it is the most important piece of information allowing to test alternative evolutionary hypothesis in many areas of research, from developmental to ecological questions.
Along with migration rates and patterns, it is fundamental to test biogeographical hypothesis-->
Coupled to species number and distribution, it is the basic information for the study of diversification processes, such as the tempo and mode of speciation and extinction, crucial for the understanding of how biodiversity patterns are shaped across space and time [@Morlon2014].

When organisms are preserved in a fossil form, a time frame of taxon origin can be obtained directly from the age of rock strata.
However, not all organisms fossilize well or at all.
Fossilization success alone is highly circumstancial, and varies depending on a number of parameters including the nature of the habitat, population size, species range breadth and physical characteristics of the organism. <!--fossil recovery is also highly circumstancial-->
Thus, relying only on the fossil record to obtain a time frame of lineage divergence<!--taxon ages--> for all life is<!--, alas,--> not possible.
Relative rate of DNA or aminoacid substitution constitutes another important source of information on lineage divergence. <!--describe the two components of rel subst rate: which, contain tne information on absolute rate of substitution and absolute time.-->
It is usually obtained from hypothesis of character homology (alignments) when reconstructing phylogenetic relationships. Molecular dating techniques use external data such as absolute time calibrations (e.g., fossils, geologic events) or absolute substitution rates <!--birds--> to generate dated phylogenies (chronograms) which contain information on absolute times of node divergence and taxon ages. <!--With molecular dating methods, an external calibration point is still necessary to estimate the absolute time of origin. Hence, the accumulation of data on fossils for calibration is also continual-->
<!--This method has been particularly useful to obtain time frame of origin of organisms with poor or no fossil record, but is also largely used for organisms with rich fossil records-->
<!--Models allow to take into account the nature of the fossil record: Given the characteristics of the fossilization process, the quality of dates inferred directly from it is not always considered the best.-->

@@

<!--What led to datelife development?
Describe public and research necessities covered by datelife
It is important to use available data on time frame of lineage origin:
To know the state of dating for a group of interest:
What range of estimated ages exist already? Are fossil and molecular
time frames coherent? (e.g., [@Magallon2015c]).
To construct a time tree of life.
For science communication, improve scientific discussions, time-framing
other events of importance in other research areas.
Comparative method: autocorrelation (phylogenetic relatedeness) corrections, phylogenetic signal
-->
Despite its importance, analytical tools to summarize available information on taxon ages are still lacking.
Data might not have been exploited because:
Data is in different repositories and formats;
Lineage names are different among studies and difficult to reconcile;
Taxonomy is also different among studies and difficult to reconcile.
<!--Data curation is necessary at some point (is this always true?)-->
Also, data curation is an important part of any biological study. The research community considers it as an important or even crucial step before data analysis. Hence, automated processes for large data analysis are frequently received with skepticism.
Recent work on this area (e.g., supersmart<!--and, which others?-->) aims to:
Generate new dates using all available DNA sequence information;
Perform one global analysis using all available information;
Problems or downsides: This might be time consuming for large groups and a lot of data curation is still necessary.
<!--General issues with dating techniques:-->
Choosing correct fossils for calibration requires a lot of expertise and knowledge on the group.
Incorrect use of fossils can generate severe bias in dating results [@Sauquet2012c].
DateLife palliates this by using only information available from already published studies, which are ideally constructed using robust information, such as sequence data and curated fossil calibrations.
<!--Data curation is largely based on taxonomic knowledge; or at least on the interplay between taxonomy and phylogenetics-->
<!--Importance of datelife:-->
DateLife can summarize this information in several formats that can be easily inspected by users.
This  allows rapidly obtaining a time frame of taxon divergence for a wide number of taxa.
DateLife can also generate chronograms for taxa with little available information, by using the available data as calibration points.
DateLife is the main service for scaling phylogenetic trees in Phylotastic! system [@Stoltzfus2013] <!--to construct phylogenetic trees on the fly [http://phylotastic.org/].-->
It can be used through an R package<link to documentation>, a web interface (<http://www.datelife.org/query/>) and an API<!--(still not up, right?)-->.

@@

DateLife also stores information on input taxon presence/absence across subset chronograms. Users can choose to add ages of missing taxa to subset chronograms in different ways, depending on the amount of knowledge or how much they want to be involved in the steps of the addition process.
If users have no access to biological information (i.e., a character, DNA or protein matrix), missing taxa can be added to any chronogram simply at random, or by following taxonomic or phylogenetic knowledge from expert sources. There are a wide number of open reference taxonomies available, such as the Catalogue of Life [@Roskob2017] or the NCBI taxonomy database [@Federhen2012]. Expert phylogenies (with or without branch lengths) to be used as topological constraint (backbone) can also be obtained from a number of public repositories, such as OToL [@Hinchliff2015], TreeBASE [@Piel2002] and Dryad (<https://www.datadryad.org//>).
At the moment, DateLife only uses OToL's synthetic tree and reference taxonomy as expert knowledge to automatically add missing taxa to chronograms. Alternatively, users can input a reference taxonomy or topological constraint of their choosing or making. <!-- at some point we will implement the possibility for users to choose different reference taxonomies automatically-->
If OToL's synthetic tree is not satisfactorily resolved for the taxa of interest, DateLife can construct a sequence data matrix from DNA markers available from the Barcode of Life Database (BOLD; @Ratnasingham2007), to attempt to further resolve polytomies. It will follow OToL's synthetic tree as backbone. <!--This molecular tree can then be dated with chronosMPL from the ape package, but is it really needed?? To estimate node ages chronosMPL uses mean path length method from Britton et al. 2002-->
<!--It can rapidly construct trees with branch lengths from
a set of lineages with sequence data from the Barcode Of Life Database (BOLD, <http://www.boldsystems.org/>), using the synthetic OToL (<https://tree.opentreeoflife.org>) as backbone, and to obtain divergence dates.
It allows direct comparison of dates obtained with different markers
available in BOLD (in plants and fungi in particular).-->
To use information from a topological constraint, DateLife calls the congruification method described in [@Eastman2013] to find shared nodes between trees (congruent nodes). It then fixes their ages, and add ages to remaining nodes with a dating method that can be specified by the user. If users have access to biological data, thay can input a tree with branch lengths proportional to relative substitution rates as topological constraint. In this case, age data from congruent nodes will be used as calibration points.
Age data from several chronograms can be combined and congruified to be used as calibration points in a single analysis.

Several dating methods are implemented in DateLife.
Branch Length Adjuster (BLADJ) is a simple algorithm to distribute ages of undated nodes evenly, which minimizes age variance in the chronogram [@Webb2008].
DateLife implements BLADJ from the development R version of phylocom's R package [@Webb2008], phylocomr (<https://github.com/ropensci/phylocomr>). It can only be used when there is a topological constraint with no branch lengths.
PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees. It is also called through R. <!--It could be used with a reference taxonomy, if we provide a tree with politomies-->
treePL, is a semi-parametric, rate-smoothing, penalized likelihood dating method [@Smith2012]. It is called through R.
MrBayes program [@Huelsenbeck2001; @Ronquist2003] can be used when adding taxa at random, following a reference taxonomy or a topological constraint. It draws ages from a pure birth model, as implemented by Jetz and collaborators [-@Jetz2012].
DateLife calls MrBayes trough an R function.

@@

<!--Potential applications shown here-->Here, we have shown that DateLife allows an easy and fast obtention of all publicly available information on taxon ages, which can be used to generate new data. <!--Potential applications not shown here-->This information can be used to account for the effect of phylogenetic signal in studies of trait evolution; to explore potential speciation and extinction dynamics of interest within a clade; to obtain a time frame of biogeographical events; for science communication and outreach, amongst others.
<!--Why you should use it-->Compared to similar platforms such as time tree of life and supermart, it offers several advantages.
It is fast;
source data is completely open;
it requires no expert biological knowledge from users for any of its functionalities;
it allows exploration of alternative taxonomic and phylogenetic schemes;
it allows rapid exploration of the effect of alternative divergence time hypothesis;
it allows rapid synthesis in a number of different formats; <!-- we should add graphical options too-->
it facilitates reproducibility of analyses;
